Search for Spatially Extended \fermi-LAT Sources Using Two Years of Flight Data

### 2011-08-23 extended_catalog/paper/extended_catalog.tex
alexs changes

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\sr}{\text{sr}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\newcommand{\km}{\text{km}\xspace}
\newcommand{\phflux}{\ensuremath{\ph\ \cm^{-2}\s^{-1}}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{inc}}}\xspace}
\newcommand{\loglikelihood}{\ensuremath{LL}\xspace}
\newcommand{\likelihood}{\ensuremath{L}\xspace}
\newcommand{\chandra}{\text{{\em Chandra}}\xspace}
\newcommand{\swiftxrt}{\text{{\em Swift}/XRT}\xspace}
\newcommand{\rosat}{\text{{\em ROSAT}}\xspace}
\newcommand{\suzaku}{\text{{\em Suzaku}}\xspace}
\newcommand{\asca}{\text{{\em ASCA}}\xspace}
\newcommand{\einstein}{\text{{\em Einstein}}\xspace}
\newcommand{\xmmnewton}{\text{{\em XMM-Newton}}\xspace}
\newcommand{\fermi}{\textit{Fermi}\xspace}

@@

\usepackage{color}
\newcommand{\hl}[1]{{\color{red}#1}}

@@

\author{
J.~Lande\altaffilmark{2},
S.~Funk\altaffilmark{2},
M.~Ackermann (???),
A.~Allafort\altaffilmark{2},
K.~Bechtol\altaffilmark{2},
M.-H.~Grondin\altaffilmark{53},
\ldots
\altaffiltext{2}{W. W. Hansen Experimental Physics Laboratory, Kavli Institute for Particle Astrophysics and Cosmology, Department of Physics and SLAC National Accelerator Laboratory, Stanford University, Stanford, CA 94305, USA}
\altaffiltext{53}{Institut f\"ur Astronomie und Astrophysik, Universit\"at T\"ubingen, D 72076 T\"ubingen, Germany}
}

\begin{abstract}
We present a new method for quantifying
the spatial extension of sources with
the Large Area Telescope (LAT), the primary science instrument on the
{\em \fermi Gamma-ray Space Telescope} (\fermi). 
Spatial extension
is an important characteristic for correctly 
associating LAT sources with their counterparts at other
wavelengths. We perform
a series of Monte Carlo studies to validate this tool and calculate
the LAT's threshold for detecting the spatial extension of sources.  We then
use this tool to test all sources in the two year source catalog for
extension. We report the detection of nine spatially extended sources
in addition to the twelve spatially extended sources reported in the
second \fermi-LAT Catalog (2FGL).
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the {\em \fermi
Gamma-ray Space Telescope} (\fermi) that been surveying the $\gamma$-ray
sky since 2008 June.  The LAT has broad energy coverage (20 \mev to $>300$
\gev), wide field of view ($\sim 2.4 \sr$), and large effective area
($\sim 8000\ \cm^2$ at $>1 \gev$).

Using two years of all-sky surveying data, the LAT Collaboration published
a catalog of 1873 \gev sources \citep{second_cat}. Many of these have
been positionally associated with a variety of source classes including
Active Galactic Nuclei (AGN), pulsars, normal galaxies, and supernova
remnants (SNRs).  While the counterparts of many LAT sources can be
spatially resolved when observed at other frequencies, detection of
spatial extension of these sources at \gev energies is difficult because
the size of the point-spread function (PSF) of the LAT is comparable
to the typical size of many of these sources.

However, a number of sources positionally coincident with LAT sources
exhibit extensions at other wavelengths which are comparable to or larger than 
the LAT PSF.  In particular, certain SNRs, pulsar wind nebulae (PWNe),
molecular clouds, normal galaxies, galaxy clusters or dark matter
satellites could be expected to be detected as spatially extended
sources by the LAT.  
The LAT Collaboration has previously reported on five SNRs which are
spatially extended beyond the LAT PSF at \gev energies 
(IC443, W28, W44,
W51C, and RX J1713.7-3946 \citep{ic443,w28,w44,w51c,rx_j1713_lat}). In
addition, three extended PWNe were detected: MSH 15-52, Vela~X,
and HESS\,J1825-137 \citep{msh1552,velax,fermi_hess_j1825}, as well
as two close-by galaxies, the Large and Small Magalenic
Cloud \citep{lmc,smc}, and finally the radio galaxy Centarus A
\citep{cen_a_lat}.
For each of these sources, the extension was determined from a dedicated
analysis of the particular source and so a systematic
scan of all LAT detected sources could uncover
additonal spatially extended sources.

\hl{
Many extended sources in the Galactic plane have been detected at \tev
energies using air Cherenkov detectors.  Most prominent was a survey of
the Galactic plane using the High Energy Stereoscopic System (H.E.S.S)
which reported 14 spatially extended sources with extensions varying from
$\sim0\fdg1$ to $\sim0\fdg25$ \citep{hess_plane_survey}.  In fact, within
our Galaxy only very few sources (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61-303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes})
have no significant extension detected at \tev energies.  High energy
$\gamma$-rays from these sources are produced by the decay of $\pi^0$s
produced by hadronic interactions with interstellar matter and by
relativistic electrons due to Inverse Compton (IC) scattering and
Bremsstrahlung radiation \citep{blandford_and_eichler_1987}.  It is likely
that the \gev and \tev emission from these sources originates from the
same population of high-energy particles and so at least some of these
sources should be detected as extended sources at \gev energies. By
studying the spectra of these \tev sources at \gev energies, we may help
to determine the emission mechanisms of these high energy sources.
}

The capability to spatially resolve \gev $\gamma$-ray sources is important
for several reasons.  First, multiple potential counterparts detected at
other frequencies exist within the localization confidence region of a
typical LAT source along the Galactic plane.  Finding a coherent source
extension across different energy bands can help to associate a
LAT source to an otherwise confused counterpart.  Also, due to the strong
energy dependence of the LAT PSF, the spatial and spectral information
about a source do not simply decouple. An inaccurate spatial model
will bias the spectral model of the source. Specifically, 
modeling a spatially extended source as point-like will systematically
shift a spectral analysis to softer indices. Furthermore,
correctly modeling a source's extension is important for improving
the model of the sky and removing excess residuals, for example in the
region around surrounding the Large Magellanic Clouds \citep{first_cat}.
Such excess residuals potentially bias the significance
and measured spectrum of neighboring sources in the densely populated
Galactic plane.

\section{Analysis Methods}

Morphological studies of sources using the LAT are challenging
because of the strongly energy-dependent PSF that is comparable in
size to the extension of many sources expected to be detected at
\gev energies.  Additional complications arise for sources along
the Galactic plane due to systematic uncertainties in the Galactic
diffuse emission.  The LAT's PSF is limited at its lowest detectable
energies by multiple scattering in the silicon strip tracking section
of the detector and is several degrees at 100 \mev.  The PSF improves
with energy approaching a 68\% containment radius of $\sim0\fdg2$ at
the highest energies (when averaged over the acceptance of the LAT)
and is limited by the granularity of the silicon strips in the tracker
\citep{atwood_LAT_mission,on_orbit_calibration,lat_on_orbit_psf}.\footnote{More
information about the performance of the LAT can be found at the Fermi
Science Support Center (FSSC,\url{http://fermi.gsfc.nasa.gov}).} However,
since most high energy astrophysical sources have spectra that decrease
rapidly with increasing energy, the improved resolution of the higher
energy photons is offset by their low statistics. Therefore sophisticated
analysis techniques are required to maximize our sensitivity
to these extended sources.

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT. The tool performs
a maximum likelihood analysis in which the Poisson likelihood for
observing the measured counts is maximized, given a parametrized
spatial and spectral model of the source and its surrounding region.
The extension of a source can be modeled by a geometric shape (e.g. a
disk or Gaussian) and the source's position, extension, and spectrum
can be simultaneously fit.

This type of analysis is not feasible using the standard LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
FSSC.} because \gtlike can only
fit the spectral parameters of the model unless a more sophisticated
iterative procedure is used to also test various source morphologies.
We note that \gtlike has been used in the past in several studies of
source extension in the LAT Collaboration \citep{lmc,smc,w28,w51c}.
Based upon a set of \gtlike maximum likelihood fits at fixed extensions,
a profile of the likelihood as a function of the extension was calculated.
This approach is not optimal because the position, extension, and
spectrum of the source must be simultaneously fit to find the best fit
parameters and to correctly compute the statistical significance of the
detection.  Furthermore since the \gtlike likelihood profile approach
is computationally intensive, no large-scale Monte Carlo effort has been
run to validate it.

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT Collaboration called \pointlike
\citep{first_cat,matthew_kerr_thesis}.  We extended the code to allow
a simultaneous fit of the source extension together with the position
and the spectral parameters.

The choice to base the spatial extension fitting on \pointlike rather than
\gtlike was made on considerations of computing time.  The \pointlike
algorithm was optimized for speed to handle larger numbers of sources
efficiently which is important for our catalog scan as well as the Monte
Carlo validation of our method.  Details on the \pointlike package can
be found in \cite{matthew_kerr_thesis}.

@@

In \pointlike, it is assumed that the spatial and spectral
model of an extended source are separable, i.e. that the source
model $M(l,b,E)=S(l,b)\times X(E)$ where $S(l,b)$ is the spatial
distribution and $X(E)$ is the spectral distribution.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a
function of energy) and uses the \minuit library to maximize the
likelihood by simultaneously varying the position and extension of
the source \citep{minuit_documentation}.  As will be described in
Section~\ref{monte_carlo_validation}, simultaneously fitting the position
and extension is important to correctly calculate the statistical
significance of the detection of extension.  For each position and
extension, the spectral parameters of the sky model are refit.  To avoid
projection effects, the fitted source position parameters are not the
source's longitude and latitude but instead the source's displacement
in a rotated reference frame.

The significance of the extension of a source can be calculated
from \tsext, which 
is the test statistic of the likelihood ratio test of
a model with a spatially extended source and a model
with a point-like source and is defined as
\begin{equation}
  \tsext=2\log(\likelihood_\text{ext}/\likelihood_\text{ps}) = 2 (\loglikelihood_\text{ext}-\loglikelihood_\text{ps})
\end{equation}
where \loglikelihood is the log of the Poisson likelihood.
\pointlike calculates \tsext by fitting a source first with a spatially
extended model and then as a point source.  The interpretation
of \tsext in terms of a statistical significance is discussed in
Section~\ref{monte_carlo_validation}.

For extended sources with an assumed radially symmetric shape,
we use a semi-analytic calculation to speed up the calculation of
the expected photon distribution. The expected photon 
distribution can be written as
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') r' dr' d\phi'
\end{equation}
For the LAT, the PSF can be parameterized by a {\em King function}
\begin{equation}
  \text{PSF}(r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $u=(r/\sigma)^2/2$ and $\sigma$ and $\gamma$ are free parameters
\citep{matthew_kerr_thesis}.  For radially symmetric extended sources,
the angular part of the integral can be evaluated analytically
\begin{align}
  \text{PDF}(u) & = \int_0^\infty r' dr'
  I_\text{src}(v) 
  \int_0^{2\pi} d\phi' 
  \text{PSF}(\sqrt{2\sigma^2(u+v-2\sqrt{uv}\cos(\phi-\phi'))})
  \\
  & = \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times {}_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{align}
where $v=(r'/\sigma)^2/2$ and ${}_2F_1$ is the Gaussian hypergeometric
function.  This convolution formula reduces the expected photon
distribution to a single numerical integral.

There will always be a small numerical discrepancy between the expected
photon distribution derived from a true point source and a very small
extended source due to numerical error in the convolution.  In most
situations, this error is insignificant.  But in particular for
very bright sources, this numerical error has the potential to bias the
test statistic for the extension test. Therefore, when calculating
\tsext, we compare the likelihood fitting the source with an extended
spatial model to the likelihood when fixing its extension to ${10^{-10}}\deg$.

We estimate the error on the extension of a source by fixing the position
of the source and varying the extension until the likelihood has fallen
by \onehalf, corresponding to a $1\sigma$ error.  This method is
shown schematically in Figure~\ref{four_plots_ic443}(a) which shows the
change in \loglikelihood when varying the extension of the SNR IC443.
The localization error is calculated by fixing the extension and spectrum
of the source and fitting to the likelihood function a Gaussian as a function
of position.

@@

\pointlike is important for LAT analyses that require many iterations
such as source localization and extension fitting.  On the other hand,
because of the fewer approximations used in calculating the likelihood
we expect the spectral parameters found with \gtlike to be slightly
more precise.  Furthermore, because \gtlike is the standard likelihood
analysis package, it has been more extensively validated for spectral
analysis.  For those reasons, in the following analysis  we use \pointlike
to determinate the position and extension of a source and subsequently determine
the spectrum using \gtlike. Both \gtlike and \pointlike can be 
used to estimate
the statistical significance of the extension of a source and we require
that both methods agree to claim a detection.

We found good agreement between the two methods.  
Unless explicitly mentioned, all \ts, \tsext, and spectral
parameters were calculated using \gtlike using the best-fit
positions and extensiosn found with \pointlike.

@@

There is a degeneracy between a spatially extended source and multiple
point sources that are separated by angular distances comparable or
smaller than the size of the LAT PSF.  To assess the possibility of source
confusion, we developed a function in \pointlike to simultaneously fit
the position of two point sources.

We define \tsinc as twice the increase in \loglikelihood fitting the
region as two point sources compared to fitting the region as one point
source: 
\begin{equation}
  \tsinc=2\times(\loglikelihood_\text{2pts}-\loglikelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
or $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Similar to the case of extended sources described in Section~\ref{gtlike_crosscheck},
the spectrum of the two point sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

We tested two different models for the
surface brightness profile, either a two dimensional Gaussian
\begin{equation}\label{gauss_pdf}
  I(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}\label{disk_pdf}
  I(x,y)=
  \begin{cases}
    \frac{1}{\pi\sigma^2} & x^2+y^2\le\sigma^2 \\
    0                      & x^2+y^2>\sigma^2
  \end{cases}
\end{equation}
Although these shapes are significantly different, their convolution
with the PSF look similar.  Figure~\ref{compare_disk_gauss}
demonstrates this by showing $I_\text{src}$, the PSF, and the PDF for
a radially symmetric uniform brightness and the Gaussian surface brightness of size 0\fdg5.\footnote{To
allow a valid comparison between Gaussian and disk shaped morphologies,
we define the source size for this test as the radius containing
68\% of the intensity ($\rsixeight$).  $\rsixeight_\text{,Gaussian}=1.51\sigma$
and $\rsixeight_\text{,disk}=0.82\sigma$ where $\sigma$
is defined in Equation~\ref{gauss_pdf} and
Equation~\ref{disk_pdf} respectively.} This plot shows that the LAT is 
not very sensitive to probing the exact structure of an extended source.
In the search for extended sources, we therefore use only a uniform
disk as our spatial extension shape. We then quote the radius of the
disk edge as the size of the source.

@@

We calculated the LAT's detection threshold to resolve spatially extended
sources as being extended. We define the detection threshold as the flux at which the value
of $\tsext$ averaged over many statistical realizations of a source
is $\langle\tsext\rangle=16$, corresponding to a $4\sigma$ detection
(see section~\ref{monte_carlo_validation}).

>>>>

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\sr}{\text{sr}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\newcommand{\km}{\text{km}\xspace}
\newcommand{\phflux}{\ensuremath{\ph\ \cm^{-2}\s^{-1}}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{inc}}}\xspace}

\newcommand{\likelihood}{\ensuremath{\mathcal{L}}\xspace}

\newcommand{\chandra}{\text{{\em Chandra}}\xspace}
\newcommand{\swiftxrt}{\text{{\em Swift}/XRT}\xspace}
\newcommand{\rosat}{\text{{\em ROSAT}}\xspace}
\newcommand{\suzaku}{\text{{\em Suzaku}}\xspace}
\newcommand{\asca}{\text{{\em ASCA}}\xspace}
\newcommand{\einstein}{\text{{\em Einstein}}\xspace}
\newcommand{\xmmnewton}{\text{{\em XMM-Newton}}\xspace}
\newcommand{\fermi}{\textit{Fermi}\xspace}

@@

\usepackage{color}

%\newcommand{\hl}[1]{{\color{red}#1}}
\newcommand{\hl}[1]{#1}

@@

\author{
J.~Lande\altaffilmark{2},
S.~Funk\altaffilmark{2},
M.~Ackermann (???),
A.~Allafort\altaffilmark{2},
M.-H.~Grondin\altaffilmark{53},
% Keith Bechtol
% Marshall Roth?
% Marianne Lemoine-Goumard
% Francesco Giordano
% Toby Burnett
% Jean Ballet
% Matthew Kerr
\ldots
\altaffiltext{2}{W. W. Hansen Experimental Physics Laboratory, Kavli Institute for Particle Astrophysics and Cosmology, Department of Physics and SLAC National Accelerator Laboratory, Stanford University, Stanford, CA 94305, USA}
\altaffiltext{53}{Institut f\"ur Astronomie und Astrophysik, Universit\"at T\"ubingen, D 72076 T\"ubingen, Germany}
}

\begin{abstract}
We present a new method for quantifying the spatial extension of sources
with the Large Area Telescope (LAT), the primary science instrument on
the {\em \fermi Gamma-ray Space Telescope} (\fermi).  Spatial extension
is an important characteristic for correctly associating LAT sources with
their counterparts at other wavelengths. We perform a series of Monte
Carlo simulations to validate this tool and calculate the LAT's threshold
for detecting the spatial extension of sources.  We then use this tool
to test all sources in the two-year source catalog for extension. We
report the detection of nine spatially extended sources in addition to
the twelve spatially extended sources reported in the second \fermi-LAT
Catalog (2FGL).
\end{abstract}

@@

\hl{

A number of astrophysical sources classes including supernova remnant
(SNRs), pulsar wind nebulae (PWNe), molecular clouds, normal galaxies,
galaxy clusters could be expected to spatially resolvable at \gev
energies.  Dark matter satellites are also hypothesized to be seen
as spatially extended at \gev energies.  The LAT Collaboration has
previously reported on five SNRs which are spatially extended beyond
the LAT PSF at \gev energies (IC443, W28, W44, W51C, and RX J1713.7-3946
\citep{ic443,w28,w44,w51c,rx_j1713_lat}). In addition, three extended PWNe
were detected as being extended: MSH 15-52, Vela~X, and HESS\,J1825-137
\citep{msh1552,velax,fermi_hess_j1825}. Two close-by galaxies the Large
and Small Magalenic Clouds and one radio galaxy Centarus A were spatially
resolved at \gev energies \citep{lmc,smc,cen_a_lat}.  Furthermore,
a number of other sources previously detected at \gev energies are
positionally coincident with sources that exhibit extension at other
wavelengths and could be expected to be spatially resolvable at \gev
energies.

At \tev energies many sources have been recently detected as
begin spatially extended using air Cherenkov detectors.  Most
prominent was a survey of the Galactic plane using the High Energy
Stereoscopic System (H.E.S.S) which reported 14 spatially extended
sources with extensions varying from $\sim0\fdg1$ to $\sim0\fdg25$
\citep{hess_plane_survey}.  In fact, within our Galaxy only very few
sources detected at \tev energies (most notably the $\gamma$-ray binaries
LS\,5039 \citep{HESSLS5039}, LS I+61-303 \citep{MAGICLSI, VERITASLSI}
HESS\,J0632+057 \citep{HESS0632}, and the Crab nebula \citep{crab_weekes})
have no significant extension detected.  High energy $\gamma$-rays from
these sources are produced by the decay of $\pi^0$s produced by hadronic
interactions with interstellar matter and by relativistic electrons
due to Inverse Compton (IC) scattering and Bremsstrahlung radiation
\citep{blandford_and_eichler_1987}.  It is likely that the \gev and
\tev emission from these sources originates from the same population of
high-energy particles and so at least some of these sources should be
detected as extended sources at \gev energies.  Studying 
these \tev sources at \gev energies would help to
determine the emission mechanisms producing these high energy photons.

The Large Area Telescope (LAT) is a pair conversion telescope on the
{\em \fermi Gamma-ray Space Telescope} (\fermi) that been surveying the
$\gamma$-ray sky since 2008 June.  The LAT has broad energy coverage
(20 \mev to $>300$ \gev), wide field of view ($\sim 2.4 \sr$), and large
effective area ($\sim 8000\ \cm^2$ at $>1 \gev$).  Using two years of
all-sky surveying data, the LAT Collaboration published a catalog of 1873
sources detected at these \gev energies called 2FGL \citep{second_cat}.
The counterparts of many of these sources can be spatially resolved
when observed at other frequencies but detecting the spatial extension
of these sources at \gev energies is difficult because the size of the
point-spread function (PSF) of the LAT is comparable to the typical size
of many of these sources.

On the other hand, the capability to spatially resolve \gev $\gamma$-ray
sources is important for several reasons.  First, multiple potential
counterparts detected at other frequencies exist within the localization
confidence region of a typical LAT source along the Galactic plane.
Finding a coherent source extension across different energy bands can
help to associate a LAT source to an otherwise confused counterpart.
Furthermore, dark matter satellite in our galaxy have been predicted to be
detected and spatially extended at \gev energies and could be discovered
outside the galactic plane \citep{pre_luanch_dark_matter_fermi}.  Also,
due to the strong energy dependence of the LAT PSF, the spatial and
spectral information about a source do not simply decouple. An inaccurate
spatial model will bias the spectral model of the source. Specifically,
modeling a spatially extended source as point-like will systematically
shift a spectral analysis to softer indices. Furthermore, correctly
modeling a source's extension is important for improving the model of
the sky and removing excess residuals, for example in the region around
surrounding the Large Magellanic Clouds \citep{first_cat}.  Such excess
residuals potentially bias the significance and measured spectrum of
neighboring sources in the densely populated Galactic plane.

Each of the previously detected extended LAT sources was spatially
resolved by a dedicated analysis of the particular source so
we expect that a systematic scan of all LAT detected sources
could uncover additional spatially extended sources.  For these
reasons, in Section~\ref{analysis_methods_section} we present
a new method for analyzing spatially extended LAT sources. In
Section~\ref{monte_carlo_validation} we demonstrate that this tool
can be used for an unbiased statistical test of the extension of
LAT sources and in Section~\ref{extension_sensitivity} we calculate
the LAT's detection threshold to resolve the extension of a source.
In Section~\ref{extended_source_search_method} we describe a search for
new spatially extended LAT sources and in Section~\ref{validate_known}
we reanalyzing the twelve previously detected extended sources
included in 2FGL.  In Section~\ref{test_2lac_sources} we further
validate our extension test by testing the extension of Active
galactic nucleus (AGN) believed to be unresolvable. Finally, in
Section~\ref{new_ext_srcs_section} we present the detection of the
extension of nine spatially extended sources that were reported in 2FGL
but not previously spatially resolved.

\section{Analysis Methods}
\label{analysis_methods_section}

Morphological studies of sources using the LAT are challenging
because of the strongly energy-dependent PSF that is comparable in
size to the extension of many sources expected to be detected at
\gev energies.  Additional complications arise for sources along
the Galactic plane due to systematic uncertainties in the Galactic
diffuse emission.  The LAT's PSF is limited at its lowest detectable
energies by multiple scattering in the silicon strip tracking section
of the detector and is several degrees at 100 \mev.  The PSF improves
with energy approaching a 68\% containment radius of $\sim0\fdg2$ at
the highest energies (when averaged over the acceptance of the LAT)
and is limited by the granularity of the silicon strips in the tracker
\citep{atwood_LAT_mission,on_orbit_calibration,lat_on_orbit_psf}.\footnote{More
information about the performance of the LAT can be found at the \fermi
Science Support Center (FSSC,\url{http://fermi.gsfc.nasa.gov}).} However,
since most high energy astrophysical sources have spectra that decrease
rapidly with increasing energy, the improved resolution of the higher
energy photons is offset by their low statistics. Therefore sophisticated
analysis techniques are required to maximize our sensitivity
to these extended sources.

@@

A new analysis tool has been developed to address the unique requirements
for studying spatially extended sources with the LAT. The tool performs
a maximum likelihood analysis in which the Poisson likelihood to find
the observing counts is maximized given a parametrized spatial and
spectral model of the source and its surrounding region.  The sky is
divided into cubes in space and energy using the healpix representation
of the sky \citep{healpix} and the likelihood is maximized over all bins
in a region.  The extension of a source can be modeled by a geometric
shape (e.g. a disk or a two-dimensional Gaussian) and the source's position, extension,
and spectrum can be simultaneously fit.

This type of analysis is not feasible using the standard LAT likelihood
analysis tool \gtlike\footnote{\gtlike is distributed publicly by the
FSSC.} because it can only fit the spectral parameters of the model
unless a more sophisticated iterative procedure is used to also test
various source morphologies.  We note that \gtlike has been used in the
past in several studies of source extension in the LAT Collaboration
\citep{lmc,smc,w28,w51c}.  In these studies, a profile based upon a
set of \gtlike maximum likelihood fits at fixed extensions was used
to build a profile of the likelihood as a function of extension.
This approach is not optimal because the position, extension, and
spectrum of the source must be simultaneously fit to find the best fit
parameters and to correctly compute the statistical significance of
a detection.  Furthermore since the \gtlike likelihood profile approach
is computationally intensive, no large-scale Monte Carlo simulations
have been run to validate it.

The approach presented here is based on a second maximum likelihood
fitting package developed in the LAT Collaboration called \pointlike
\citep{first_cat,matthew_kerr_thesis}.  The choice to base the
spatial extension fitting on \pointlike rather than \gtlike was made
on considerations of computing time.  The \pointlike algorithm was
optimized for speed to handle larger numbers of sources efficiently
which is important for our catalog scan and for being able
to perform large-scale Monte Carlo simulations to validate the tool.
Details on the \pointlike package can be
found in \cite{matthew_kerr_thesis}.  We extended the code to allow a
simultaneous fit of the source extension together with the position and
the spectral parameters.

@@

In \pointlike, it is assumed that the spatial and spectral
model of an extended source are separable, i.e. that the source
model $M(l,b,E)=S(l,b)\times X(E)$ where $S(l,b)$ is the spatial
distribution and $X(E)$ is the spectral distribution.  To fit an extended source,
\pointlike convolves the extended source shape with the PSF (as a
function of energy) and uses the \minuit library 
\citep{minuit_documentation}
to maximize the
likelihood by simultaneously varying the position and extension of
the source.  As will be described in
Section~\ref{monte_carlo_validation}, simultaneously fitting the position
and extension is important to correctly calculate the statistical
significance of the detection of extension.  For each position and
extension, the spectral parameters of the sky model are refit.  To avoid
projection effects, the fitted source position parameters are not the
source's longitude and latitude but instead the source's displacement
in a rotated reference frame.

The significance of the extension of a source can be calculated from the
likelihood ratio test of a model with a spatially extended source and
a model with a point-like source. The test statistic for this procedure
is defined as
\begin{equation}
  \tsext=2\log(\likelihood_\text{ext}/\likelihood_\text{ps}) 
\end{equation}
where \likelihood is the Poisson likelihood.
\pointlike calculates \tsext by fitting a source first with a spatially
extended model and then as a point source.  The interpretation
of \tsext in terms of a statistical significance is discussed in
Section~\ref{monte_carlo_validation}.

For extended sources with an assumed radially symmetric shape,
we can optimize the calculation by performing one
of the integrals analytically.
The expected photon 
distribution can be written as
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') r' dr' d\phi'
\end{equation}
For the LAT, the PSF can be parameterized by a King function \citep{king_function}:
\begin{equation}
  \text{PSF}(r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma},
\end{equation}
where $u=(r/\sigma)^2/2$ and $\sigma$ and $\gamma$ are free parameters
\citep{matthew_kerr_thesis}.  For radially symmetric extended sources,
the angular part of the integral can be evaluated analytically
\begin{align}
  \text{PDF}(u) & = \int_0^\infty r' dr'
  I_\text{src}(v) 
  \int_0^{2\pi} d\phi' 
  \text{PSF}(\sqrt{2\sigma^2(u+v-2\sqrt{uv}\cos(\phi-\phi'))})
  \\
  & = \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma 
  \times {}_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right).
\end{align}
where $v=(r'/\sigma)^2/2$ and ${}_2F_1$ is the Gaussian hypergeometric
function.  This convolution formula reduces the expected photon
distribution to a single numerical integral.

There will always be a small numerical discrepancy between the expected
photon distribution derived from a true point source and a very small
extended source due to numerical error in the convolution.  In most
situations, this error is insignificant.  But in particular for
very bright sources, this numerical error has the potential to bias the
test statistic for the extension test. Therefore, when calculating
\tsext, we compare the likelihood fitting the source with an extended
spatial model to the likelihood when fixing its extension to $10^{-10\,\circ}$.

We estimate the error on the extension of a source by fixing
the position of the source and varying the extension until the
likelihood has fallen by \onehalf, corresponding to a $1\sigma$ error
\citep{Statistical_methods_book}.  This method is shown schematically in
Figure~\ref{four_plots_ic443} which shows the change in the log of the
likelihood when varying the extension of the SNR IC443.  The localization
error is calculated by fixing the extension and spectrum of the source
and fitting to the likelihood function a 2D Gaussian as a function
of position.

@@

\pointlike is important for LAT analyses that require many iterations
such as source localization and extension fitting.  On the other hand,
because \gtlike makes fewer approximations in calculating the likelihood
we expect the spectral parameters found with \gtlike to be slightly more
precise.  Furthermore, because \gtlike is the standard likelihood analysis
package, it has been more extensively validated for spectral analysis.
For those reasons, in the following analysis we used \pointlike to
determine the position and extension of a source and subsequently derived
the spectrum using \gtlike. Both \gtlike and \pointlike can be used to
estimate the statistical significance of the extension of a source and we
required that both methods agree for a source to be considered extended.
We found good agreement between the two methods.  Unless explicitly
mentioned, all \ts, \tsext, and spectral parameters were calculated using
\gtlike with the best-fit positions and extension found by \pointlike.

@@

There is a degeneracy between a spatially extended source and multiple
point sources separated by angular distances comparable to or smaller than
the size of the LAT PSF.  To assess the possibility of source confusion,
we use \pointlike to simultaneously fit the position of two point sources
in the region of extended source candidates.

We define \tsinc as twice the increase in the log of the likelihood fitting the
region as two point sources compared to fitting the region as one point
source: 
\begin{equation}
  \tsinc=2\log(\likelihood_\text{2pts}/\likelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
or $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Similar to the case of extended sources described in Section~\ref{gtlike_crosscheck},
the spectra of the two point sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

We tested two different models for the
surface brightness profile, either a 2D Gaussian
\begin{equation}\label{gauss_pdf}
  I(x,y)=\tfrac{1}{2\pi\sigma^2}\exp\left(-(x^2+y^2)/2\sigma^2\right)
\end{equation}
or a uniform disk
\begin{equation}\label{disk_pdf}
  I(x,y)=
  \begin{cases}
    \frac{1}{\pi\sigma^2} & x^2+y^2\le\sigma^2 \\
    0                      & x^2+y^2>\sigma^2
  \end{cases}
\end{equation}
Although these shapes are significantly different, when convolved
with the PSF, their profiles are very similar.  
This is demonstrated in 
Figure~\ref{compare_disk_gauss}
by showing $I_\text{src}$, the PSF, and the PDF for
a radially symmetric uniform brightness and a Gaussian surface brightness of size 0\fdg5.\footnote{To
allow a valid comparison between Gaussian and disk shaped morphologies,
we define the source size for this test as the radius containing
68\% of the intensity ($\rsixeight$).  $\rsixeight_\text{,Gaussian}=1.51\sigma$
and $\rsixeight_\text{,disk}=0.82\sigma$ where $\sigma$
is defined in Equation~\ref{gauss_pdf} and
Equation~\ref{disk_pdf} respectively.} This plot shows that the LAT 
has little sensitivity to
the exact structure of an extended source.
Therefore, in our
search for extended sources, we use only a uniform
disk as our spatial extension shape. We then quote the radius of the
disk edge as the size of the source.

@@

We calculated the LAT's detection threshold to detect that a spatially extended
sources is extended. We define the detection threshold as the flux at which the value
of $\tsext$ averaged over many statistical realizations of a source
is $\langle\tsext\rangle=16$, corresponding to a $4\sigma$ detection
(see section~\ref{monte_carlo_validation}).
